Design and implementaiton of a live coding environment for data science

### 2017-10-26 live/paper.tex
Updating informal gamma intro

\usepackage{lineno,hyperref,xcolor}
\usepackage{stmaryrd}
\usepackage{amssymb}
\usepackage{xypic}
\usepackage{semantic}
\usepackage{booktabs} 
\usepackage{subcaption}

@@

In this paper, we describe a live programming environment for data science that provides instant 
previews and contextual hints, while allowing the user to edit code in an unrestricted way in
a text editor. This is challenging as the user can make arbitrary edits and we cannot fully 
recompute previews after every change. The presented interpreter partially evalautes programs as 
they are written and reuses results of previous evaluations, making it possible to provide instant 
feedback during development.

@@

Increasing number of programming environments aim to provide the same live experience for more
standard programming languages, but doing this is not easy. Fully recomputing the whole program after
every single change is inefficient and calculating how a change in source code changes the result
is extremely hard when the editor allows arbitrary manipulation of program text. For example, 
consider the following simple program that gets the release years of 10 most expensive movies 
in a data set \ident{movies}:
%
\begin{equation*}
\begin{array}{l}  
\kvd{let}~\ident{top} =\ident{movies}\\
\quad .\ident{sortBy}(\lambda x \rightarrow x.\ident{getBudget}())\ident{.take}(\num{10})\\
\quad .\ident{map}(\lambda x \rightarrow x.\ident{getReleased()}.\ident{format}(\str{yyyy}))}\\
\end{array}
\end{equation*}
%
A live coding environment shows a preview of the list of dates \ident{top} and the programmer then
modifies the code by making the constant $\num{10}$ a variable and changing the date format to see 
the full date:
%
\begin{equation*}
\begin{array}{l}  
\kvd{let}~\ident{count} = \num{10}\\
\kvd{let}~\ident{top} = \ident{movies}\\
\quad .\ident{sortBy}(\lambda x \rightarrow x.\ident{getBudget}())\ident{.take}(\ident{count})\\
\quad \ident{.map}(\lambda x \rightarrow x.\ident{getReleased()}.\ident{format}(\str{dd-mm-yyyy}))}\\
\end{array}
\end{equation*}
%
Ideally, the live coding environment should understand the change, reuse a cached result of the
first two transformations (sorting and taking 10 elements) and only evaluate the last \ident{map}
to differently format the release dates of already computed top 10 movies.

This is not difficult to achieve if we represent the program in a structured way and allow the user
to edit code only via well-understood primitive operations such as ``extract variable'' (which has
no effect on the result) or ``change constant value'' (which forces recomputation of subsequent
transformations). However, many programmers prefer to edit programs as free form text.
We present the design and implementation of a live coding system that is capable
of reusing previously evaluated expressions as in the example above, yet, is integrated into an
ordinary text editor. The main contributions of this paper are:

\begin{itemize}
\item We introduce The Gamma (Section~\ref{sec:live}), a simple live coding environment for
  data science. We review its design and implementation and explain how it 
  bridges the gap between programming and spreadsheets.

@@

\item In the final more speculative part (Section~\ref{sec:design}), we consider alternative
  language designs that would enable further live coding experiences, which are difficult to
  build using our current system.
\end{itemize}

@@

\begin{figure*}
\includegraphics[scale=0.45]{scrape.png}
\caption{xxx}  
\end{figure*}

@@

The Gamma introduction and what this tries to do. See pretty screenshot.

This is what we care about - it's not live programming for your typical haskell,
but for R or Python.

\newpage
~

@@

>>>>

\usepackage{lineno,hyperref,xcolor}
\usepackage{stmaryrd}
\usepackage{amssymb}
\usepackage{xypic}
\usepackage{semantic}
\usepackage{booktabs} 
\usepackage{subcaption}
\usepackage{enumitem}
\setlist{leftmargin=6mm}

@@

In this paper, we describe a live programming environment for data science that provides instant 
previews and contextual hints, while allowing the user to edit code in an unrestricted way in
a text editor.

Using a text editor is challenging as any edit can significantly change the structure of code and
fully recomputing previews after every change is too expensive. The presented interpreter partially 
evalautes programs as they are written and reuses results of previous evaluations, making it 
possible to provide instant feedback during development.

@@

Increasing number of programming environments aim to provide the same live experience for more
standard programming languages, but doing this is not easy. Fully recomputing the whole program after
every single change is inefficient and calculating how a change in source code changes the result
is extremely hard when the editor allows arbitrary manipulation of program text. For example, 
consider the following simple program that gets the release years of 10 most expensive movies 
in a data set \ident{movies}:
%
\begin{equation*}
\begin{array}{l}  
\kvd{let}~\ident{top} =\ident{movies}\\
\quad .\ident{sortBy}(\lambda x \rightarrow x.\ident{getBudget}())\ident{.take}(\num{10})\\
\quad .\ident{map}(\lambda x \rightarrow x.\ident{getReleased()}.\ident{format}(\str{yyyy}))}\\
\end{array}
\end{equation*}
%
A live coding environment shows a preview of the list of dates. Next, assume that the programmer
modifies the code by making the constant $\num{10}$ a variable and changing the date format to see 
the full date:
%
\begin{equation*}
\begin{array}{l}  
\kvd{let}~\ident{count} = \num{10}\\
\kvd{let}~\ident{top} = \ident{movies}\\
\quad .\ident{sortBy}(\lambda x \rightarrow x.\ident{getBudget}())\ident{.take}(\ident{count})\\
\quad \ident{.map}(\lambda x \rightarrow x.\ident{getReleased()}.\ident{format}(\str{dd-mm-yyyy}))}\\
\end{array}
\end{equation*}
%
Ideally, the live coding environment should understand the change, reuse a cached result of the
first two transformations (sorting and taking 10 elements) and only evaluate the last \ident{map}
to differently format the release dates of already computed top 10 movies.

This is not difficult if we represent the program in a structured way and allow the user
to edit code only via primitive operations such as ``extract variable'' (which has
no effect on the result) or ``change constant value'' (which forces recomputation of subsequent
transformations). However, many programmers prefer to edit programs as free form text.

We present the design and implementation of a live coding system that is capable
of reusing previously evaluated expressions as in the example above, yet, is integrated into an
ordinary text editor. Our main contributions are:

\begin{itemize}[itemsep=3pt]
\item We introduce The Gamma (Section~\ref{sec:live}), a simple live coding environment for
  data science. We review its design and implementation and explain how it 
  bridges the gap between programming and spreadsheets.

@@

\item In more speculative conclusions (Section~\ref{sec:design}), we consider alternative
  language designs that would enable further live coding experiences, which are difficult to
  build using our current system.
\end{itemize}

@@

\begin{figure*}
\includegraphics[scale=0.21]{wiki.png}
\includegraphics[scale=0.21]{drop.png}
\caption{Scraping 2016 science events. Preview of the Wikipedia source page (left) and live filtering (right).}
\label{fig:thegamma}
\end{figure*}

@@

% https://cacm.acm.org/blogs/blog-cacm/169199-data-science-workflow-overview-and-challenges/fulltext

The Gamma aims to make basic data exploration accessible to non-experts, such as data journalists [X], who
need to use data transparently. As such, we focus on simple scripts that can be written using
tools such as Jupyter notebooks~[X]. Those scripts follow a typical data science workflow [X];
they acquire and reformat data, run a number of analyses using the clean data and then create 
several visualizations.

An important property of data science workflow is that we work with readily available concrete data.
Data scientists load inputs into memory and refer to it in subsequent REPL interactions. They might 
later wrap completed working code into a function (and run on multiple datasets), but 
do not start with functions. We reflect this pattern in our language.

In this section, we provide a brief overview of The Gamma, a simple live coding environment for
data science. We discuss the language, type providers and the user interface,
before focusing on the algorithms behind live previews in Section~\ref{sec:formal}.

\subsection{The Gamma scripting language}
Figure~\ref{fig:thegamma} shows The Gamma script that scrapes items from a Wikipedia page, 
collects those marked as ``Events'' and filters them. The script illustrates two aspects of the
scripting language used by The Gamma -- its structure and its use of type providers for
dot-driven data exploration.

The scripting language is not intended to be as expressive as, say, R or Python and so it
has a very simple structure -- scripts are a sequence of let-bindings (that obtain or transform 
data) or statements (that produce visualizations). This reflects the fact that we always work
with concrete data and allows us to provide previews.

The most notable limitation of The Gamma is that the scripting language does not support top-level 
functions. This is not a problem for the simple scripts we consider, but it would be an issue for
a general-purpose language. We discuss potential design for functions that would still be based
on working with concrete data in Section~\ref{sec:design}.

\subsection{Dot-driven data exploration}

As ilustrated by the second let binding in Figure~\ref{fig:thegamma}, many opertaions
in The Gamma can be expressed using member access via dot. The underlying mechanism is based
on type providers [X,Y] and the specific type provider used in this example has been described 
elsewhere [X].

Given a data source (scraped Wikipedia page), the type provider generates type with members that allow
a range of transformations of the data such as grouping, sorting and filtering. Some of the members
are based only on the schema (e.g.~\ident{\textquotesingle Item Contains\textquotesingle} or 
\ident{\textquotesingle Drop Heading 2\textquotesingle}), but some may also be generated based
on (a sample of) the dataset (e.g.~the second member in \ident{\textquotesingle Heading 2 
is\textquotesingle.\textquotesingle Events[edit]\textquotesingle}).

What matters for the purpose of this paper is 
the fact that most operations are expressed via member access matters. First, it means 
that we need to provide live previews for sub-expressions formed by a chain of member accesses.
Second, it means that the result of any member access expression depends on the instance and,
possibly, on the parameters provided for a method call.

\subsection{Direct manipulation and live previews}
The screenshot in Figure~\ref{fig:thegamma} show the editor as implemented in The Gamma. This
includes live previews (discussed in this paper), but also an editor that provides spreadsheet-like
interface for editing the data transformation script.

In our implementation, previews appears below the currently selected let binding.
The preview can be of several types such as web page (left) or data table (right). In this paper,
we describe how we evaluate scripts to obtain the resulting object and we ignore how such objects
are rendered.

The Figure~\ref{fig:thegamma} also shows a special handling of expressions constructed using the
pivot type provider. The editor recognises individual data transformations and provides a simple
user interface for adding and removing transformations and changing their parameters that changes
the source code accordingly. This aspect is not discussed in the present paper.

@@

\section{Related and future}
\label{sec:future}

Integrate with Python/R via Jupyter
